Title Goes Here

### 2021-03-13 analysis/paper/paper.Rmd
Corrige tamaño de las figuras

## Figure classes
DefineClass <- function(class.opts) {
    class.name <- deparse(substitute(class.opts))
    class.fun <- function(options) {
        class <- options[[class.name]]
        class <- class.opts[[class]]
        options[names(class)] <- class
        options
    }
    invisible(knitr::opts_hooks$set(
        setNames(list(class.fun), class.name)))
}

page.width <- 210 - 30 - 25
text.width <- page.width - 20
text.height <- page.height <- 297 - 25 - 25

fig.class <- list(
  colwidth = list(
    fig.width = 5
  ),
    pagewidth = list(
        fig.width = page.width/25.4,
        out.width = paste0(page.width, "mm"),
        # fig.align = "center",
        fig.env = "figure*",
        out.extra = " "
    )
)

DefineClass(fig.class)

@@

theme_set(theme_shceof(base_size = 12))

@@

guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
  guide_colorstrip(title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

guide_colorsteps_bottom <- function(width = 20, height = 0.5, even.steps = FALSE, ...) {
  guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
                   title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019. We restrict most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also use the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985 (Equation 5.7).

Precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap].

We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between `r knitr::combine_words(LatLabel(o3wave_lats))` for each Spring and computing the Fourier spectrum.

We explore the relationship with the Southern Annular Mode (SAM). We compute the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009]. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields [@campitelli2021]. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

@@

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units). The numbers at the top-left of each panel ate the variance explained by each EOF.

@@

```{r eof-naive, fig.cap = "(ref:eof-naive-cap)"}
eof_naive[, eof[[1]]$right, by = lev] %>%
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = hgt, fill = ..level..),
                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0)) +

  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised(guide = "none")  +
  scale_y_latitude() +
  scale_x_longitude() +
  coord_polar() +
  theme(axis.text = element_blank()) +
  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
  tagger::tag_facets(tag = "panel", tag_pool = explained, tag_suffix = "")
```

@@

```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=5, fig.height=2.5}
r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
    paste("2")
}, "", ""))

@@

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Real part in shading, imaginary part in contours", fig.width=5, fig.height=5}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=4, fig.width = 5}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

@@

```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=4, fig.width = 5}
# Check that the extended index coincides with the
# original index.
cors <- era_more %>%
  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>%
  na.omit() %>%
  .[, cor(hgt, i.hgt), by = .(PC, part)]

@@

era_more %>%
  copy() %>% 
  .[, ref := year(time) > 1979] %>% 
  .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  # Zero line + rangeframe
  geom_hline(data = ~.x[, mean(hgt), by  = .(PC, part)], 
             aes(yintercept = V1), size = 0.1, color = "gray30") +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  geom_line() +
  # geom_line(data = era_less) +
  scale_alpha_manual(values = c(1, 0.2)) +

  geom_smooth(method = "lm", color = "black") +

  scale_x_datetime(NULL, date_labels = "%Y") +
  scale_y_continuous("EOF") +

@@

Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis). There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`). There is no significant trend in any of the complex parts of EOF2.

@@

```{r define-plot_regr}
plot_regr <- function(data, which_PC, which_levs, which_var) {
  data <- data %>%
    .[ PC %in% which_PC] %>%
    .[variable == which_var]

@@

  data[lat <= -20] %>%
    # .[term == "I" & lev == 50] %>%
    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
    scale_fill_divergent(var_title, breaks = AnchorBreaks(0, exclude = 0),
                         guide = guide_colorstrip_bottom(20)) +
    geom_contour_pval(aes(z = p.value)) +

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=5.4, fig.width=5}
plot_regr(regr, "PC1", c(200, 50), "hgt") +
  # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
  #              aes(z = estimate, linetype = factor(-sign(..level..))), 
  #              global.breaks = FALSE, breaks = AnchorBreaks(0, bins = 2, exclude = 0)) +
  # scale_linetype(guide = "none") +
  NULL

@@

```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=5.4, fig.width=5}
plot_regr(regr, "PC2", c(200, 50), "hgt")

@@

```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  fig.height=5.4, fig.width=5}
plot_regr(regr, "PC1", c(200, 50), "t")
```

@@

```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=3.5, fig.width=5}
t %>% 
  .[lat %between% c(-80, -35)] %>% 
  .[, .(hgt = weighted.mean(hgt, cos(lat*pi/180)),
        o3 = weighted.mean(o3, cos(lat*pi/180)),
        t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
  melt(id.vars = c("lev", "time", "lon")) %>% 
  .[, value := Anomaly(value), by = .(time, lev, variable)] %>%
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
  .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  # .[, z := metR:::standard_atmosphere$altitude(lev*100)] %>% 
  # .[, dz1 := c(NA, diff(estimate)/diff(z)) , by = .(term, lon, variable)] %>% 
  # .[, dz2 := c(diff(estimate)/diff(z), NA) , by = .(term, lon, variable)] %>% 
  # .[, dz := (dz1 + dz2)/2] %>% 
  # .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, variable)] %>%
  # .[lev < 900] %>% 
  ggplot(aes(lon, lev)) +
  geom_contour_fill(data = ~.x[variable == "t"],
                    aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
  geom_contour_tanaka(data = ~.x[variable == "t"],
                      aes(z = estimate),
                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
  geom_contour2(data = ~.x[variable == "o3"], 
                aes(z = estimate, linetype = factor(-sign(..level..))), 
                breaks = AnchorBreaks(0, exclude = 0)) +
  geom_text_contour(data = ~.x[variable == "o3"], 
                aes(z = estimate*1e9, label = signif(..level.., 2)),
                size = 3,
                skip = 1, stroke = 0.15, stroke.colour = "white", 
                label.placement = lable_placement_min(),
                breaks = AnchorBreaks(0, exclude = 0)) +

  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(20)) + 
  scale_y_level() +
  scale_linetype(guide = "none") +
  scale_x_longitude(ticks = 60) +
  facet_grid(~term) +
  axis_labs_smol +
  coord_cartesian(clip = "off")
```

@@

```{r o3-regr, fig.cap = "(ref:o3-regr-cap)", fig.height=4, fig.width = 5}

o3_regr %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour2(data = periodic(o3_mean, lon = c(0, 360)),
                aes(z = toc_z, linetype = factor(-sign(..level..))), breaks = AnchorBreaks(0, exclude = 0), 
                bins = 5) +
  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +

  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   guide = guide_colorsteps_bottom(20)) + 
  scale_x_longitude() +
  scale_y_latitude() +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(PC~term) +
  axis_labs_smol +
  tag_facets("panel")

@@

```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=5, fig.height=3}
wave1_o3 <- o3 %>% 
  .[lat %between% o3wave_lats] %>% 
  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(time, lon)] %>% 
  .[, FitWave(toc, 1), by = .(time)]

@@

```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=5, fig.height=4}
pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
  copy() %>% 
  .[, PC := NULL] %>% 
  .[]

@@

```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=2.5, fig.height=2.5}
cut(ceof$eof[[1]], 2)$left %>% 
  .[, arg := Arg(hgt)] %>% 
  .[, w := Mod(hgt)/sum(Mod(hgt))] %>% 
  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
  ggplot(aes(arg)) +
  geom_histogram(binwidth = 90*pi/180, center = 0, color = "black", fill = "gray70") +
  annotate("text",
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
           color = "black", size = 3,
           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
  geom_rug() +
  scale_y_continuous("Number of years") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 90*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) 
```

@@

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=5}
time <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)]

@@

```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary scale.", fig.width=2.5, fig.height=2.5}
eof_filtered <- era5 %>% 
  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
  .[lev == 200] %>% 
  .[lat <= -20] %>% 
  .[season(time) == "SON"] %>% 
  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
  EOF(eof ~ lon + lat | time, n = 1, data = .)

@@

```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=5, fig.height=2.5}
pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

@@

fake_eof <- era5 %>% 
  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
  .[lev == 200] %>% 
  .[lat <= -20] %>% 
  .[season(time) == "SON"] %>% 
  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals, .(lon, lat, lev, season(time))] %>% 
  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
  fake_sam[., on = .NATURAL] %>% 
  .[, hgt := hgt + sam] %>% 
  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
  EOF(eof ~ lon + lat | time, n = 1:3, data = .)

>>>>

# ## Figure classes
# DefineClass <- function(class.opts) {
#     class.name <- deparse(substitute(class.opts))
#     class.fun <- function(options) {
#         class <- options[[class.name]]
#         class <- class.opts[[class]]
#         options[names(class)] <- class
#         options
#     }
#     invisible(knitr::opts_hooks$set(
#         setNames(list(class.fun), class.name)))
# }
# 
# page.width <- 210 - 30 - 25
# text.width <- page.width - 20
# text.height <- page.height <- 297 - 25 - 25
# 
# fig.class <- list(
#   colwidth = list(
#     fig.width = 5
#   ),
#     pagewidth = list(
#         fig.width = page.width/25.4,
#         out.width = paste0(page.width, "mm"),
#         # fig.align = "center",
#         fig.env = "figure*",
#         out.extra = " "
#     )
# )
# 
# DefineClass(fig.class)

width_column <- 3.6

@@

theme_set(theme_shceof(base_size = 10))

@@

guide_colorstrip_bottom <- function(width = 15, height = 0.5, ...) {
  guide_colorstrip(title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE, ...) {
  guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
                   title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019. We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985 (Equation 5.7). For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap].

We computed the amplitude and phase of the zonal wave 1 by averaging variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) `r knitr::combine_words(LatLabel(o3wave_lats))` for each Spring and computing the Fourier spectrum.

We computed the level-dependent Southern Annular Mode index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009]. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields [@campitelli2021]. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

@@

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units). The numbers at the top-left of each panel ate the zonal variance explained by each EOF.

@@

```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=3.7, fig.height=2.2}
eof_naive[, eof[[1]]$right, by = lev] %>%
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = hgt, fill = ..level..),
                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0), 
                    range = c(0.01, 0.2)) +

  geom_qmap(~.x[lat <= -20], size = 0.1) +
  scale_fill_divergent_discretised(guide = "none")  +
  scale_y_latitude() +
  scale_x_longitude() +
  coord_polar() +
  theme(axis.text = element_blank()) +
  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
  tagger::tag_facets("rc")  +
  theme(panel.spacing = grid::unit(-.7, "lines"))
```

@@

```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=width_column, fig.height=2.5}
r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
    paste("2")
}, "", ""))

@@

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Real part in shading, imaginary part in contours", fig.width=width_column, fig.height=width_column}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=width_column, fig.width = width_column}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

@@

```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 7, fig.env = "figure*"}
# Check that the extended index coincides with the
# original index.
cors <- era_more %>%
  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>%
  na.omit() %>%
  .[, cor(hgt, i.hgt), by = .(PC, part)]

@@

era_more %>%
  copy() %>% 
  .[, ref := year(time) > 1979] %>% 
   # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, PC)] %>%
  ggplot(aes(time, hgt*1e3)) +
  # Zero line + rangeframe
  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(PC, part)], 
             aes(yintercept = V1), size = 0.1, color = "gray30") +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  geom_line() +
  # geom_line(data = era_less) +
  scale_alpha_manual(values = c(1, 0.2)) +
  geom_smooth(method = "lm", color = "black") +
  scale_x_datetime(NULL, date_labels = "%Y") +
  scale_y_continuous("EOF") +

@@

Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex parts of EOF2. The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).

@@

```{r plot_regr}
plot_regr <- function(data, which_PC, which_levs, which_var) {
  data <- data %>%
    .[ PC %in% which_PC] %>%
    .[variable == which_var]

@@

  data[lat <= -20] %>%
    # .[term == "I" & lev == 50] %>%
    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
    scale_fill_divergent(var_title, breaks = AnchorBreaks(0, exclude = 0),
                         guide = guide_colorstrip_bottom(15)) +
    geom_contour_pval(aes(z = p.value)) +

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=1.2*width_column, fig.width=width_column, dependson="plot_regr"}
plot_regr(regr , "PC1", c(200, 50), "hgt") +
  # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
  #              aes(z = estimate, linetype = factor(-sign(..level..))), 
  #              global.breaks = FALSE, breaks = AnchorBreaks(0, bins = 2, exclude = 0)) +
  # scale_linetype(guide = "none") +
  NULL

@@

```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=1.3*width_column, fig.width=width_column, dependson="plot_regr"}
plot_regr(regr, "PC2", c(200, 50), "hgt")

@@

```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  dependson="plot_regr", fig.height=1.3*width_column, fig.width=width_column}
plot_regr(regr, "PC1", c(200, 50), "t")
```

@@

```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=2.5, fig.width=width_column}
t %>% 
  .[lat %between% c(-80, -35)] %>% 
  .[, .(hgt = weighted.mean(hgt, cos(lat*pi/180)),
        o3 = weighted.mean(o3, cos(lat*pi/180)),
        t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
  melt(id.vars = c("lev", "time", "lon")) %>% 
  .[, value := Anomaly(value), by = .(time, lev, variable)] %>%
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
  .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  # .[, z := metR:::standard_atmosphere$altitude(lev*100)] %>% 
  # .[, dz1 := c(NA, diff(estimate)/diff(z)) , by = .(term, lon, variable)] %>% 
  # .[, dz2 := c(diff(estimate)/diff(z), NA) , by = .(term, lon, variable)] %>% 
  # .[, dz := (dz1 + dz2)/2] %>% 
  # .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, variable)] %>%
  # .[lev < 900] %>% 
  ggplot(aes(lon, lev)) +
  geom_contour_fill(data = ~.x[variable == "t"],
                    aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
  geom_contour_tanaka(data = ~.x[variable == "t"],
                      aes(z = estimate),
                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
  geom_contour2(data = ~.x[variable == "o3"], 
                aes(z = estimate, linetype = factor(-sign(..level..))), 
                breaks = AnchorBreaks(0, exclude = 0)) +
  geom_text_contour(data = ~.x[variable == "o3"], 
                aes(z = estimate*1e9, label = signif(..level.., 2)),
                size = 2,
                skip = 1, stroke = 0.15, stroke.colour = "white", 
                label.placement = lable_placement_min(),
                breaks = AnchorBreaks(0, exclude = 0)) +

  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(15)) + 
  scale_y_level() +
  scale_linetype(guide = "none") +
  scale_x_longitude(ticks = 60) +
  facet_grid(~term) +
  axis_labs_smol +
  coord_cartesian(clip = "off")
```

@@

```{r o3-regr, fig.cap = "(ref:o3-regr-cap)", fig.height=3, fig.width = width_column}
o3_regr %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour2(data = periodic(o3_mean, lon = c(0, 360)),
                aes(z = toc_z, linetype = factor(-sign(..level..))), 
                breaks = AnchorBreaks(0, exclude = 0), 
                bins = 5) +
  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +

  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   guide = guide_colorsteps_bottom(15)) + 
  scale_x_longitude() +
  scale_y_latitude() +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(PC~term) +
  axis_labs_smol +
  tag_facets("panel")

@@

```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=width_column, fig.height=width_column*.7}
wave1_o3 <- o3 %>% 
  .[lat %between% o3wave_lats] %>% 
  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(time, lon)] %>% 
  .[, FitWave(toc, 1), by = .(time)]

@@

```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
  copy() %>% 
  .[, PC := NULL] %>% 
  .[]

@@

```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=width_column*.8, fig.height=2.5}
cut(ceof$eof[[1]], 2)$left %>% 
  .[, arg := Arg(hgt)] %>% 
  .[, w := Mod(hgt)/sum(Mod(hgt))] %>% 
  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
  ggplot(aes(arg)) +
  geom_histogram(binwidth = 90*pi/180, center = 0, color = "black", fill = "gray70") +
  annotate("text",
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
           color = "black", size = 3,
           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
  geom_rug() +
  scale_y_continuous("Number of years") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 90*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) 
```

@@

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
time <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)]

@@

```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary units.", fig.width=2.5, fig.height=2.5}
eof_filtered <- era5 %>% 
  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
  .[lev == 200] %>% 
  .[lat <= -20] %>% 
  .[season(time) == "SON"] %>% 
  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
  EOF(eof ~ lon + lat | time, n = 1, data = .)

@@

```{r fake_eof}
pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

@@

fake_eof <- era5 %>% 
  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
  .[lev == 200] %>% 
  .[lat <= -20] %>% 
  .[season(time) == "SON"] %>% 
  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals, .(lon, lat, lev, season(time))] %>% 
  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
  fake_sam[., on = .NATURAL] %>% 
  .[, hgt := hgt + sam] %>% 
  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
  EOF(eof ~ lon + lat | time, n = 1:3, data = .)  
```
